\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-09-19 papers/methods/paper.tex
Adding comparison to BBC

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova colour population's mean and skewness are highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. {\green A common solution to this computational cost problem is to bin the supernovae, which then produces a matrix of arbitrary size. Whilst binning data does result in some loss of information, recent works tested against simulations show that this information is not cosmologically significant \citep{Scolnic2016, Kessler2017}.}

@@

{\green 
The diverse approaches and implementations applied to supernova cosmology are a response to the significant challenges and complications faced by when performing supernova cosmology analyses. In this Section, we outline several of the most prominent challenges.

@@

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SNIa, datasets are commonly formed from the discoveries of multiple surveys. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for. 
}

@@

\begin{table}
	\centering
	\caption{{\green Parameters defined in our model and a summary of their use. The parameters are broken into multiple sections, the top for parameters which are defined globally across all surveys, the second section for parameters which are defined for each survey, the third section are parameters which are defined for each supernova. The bottom section shows the observables (not parameters) that are the input data to the model.}}
	\label{tab:param_summary}
	\begin{tabular}{ll}
		\hline
		Parameter & Description \\
		\hline
		$\Omega_m$  & Matter density  \\
		$w$  &  Dark energy equation of state  \\
		$\alpha$ & Stretch standardisation   \\
		$\beta$         &  Colour standardisation   \\
		$\delta(0)$ & Scale of the mass-magnitude correction\\
		$\delta(\infty)/\delta(0)$ & Redshift-dependence of mass-magnitude correction\\
		$\delta\mathcal{Z}_i$ & Systematics scale\\
		$\delta S$ & Selection effect deviation \\
		$\langle M_B \rangle$ & Mean absolute magnitude \\
		%$s_c$ & Colour bias\\
		\hline
		$\langle x_1^i \rangle$ & Mean stretch nodes\\
		$\langle c^i \rangle$ & Mean colour nodes\\
		$\alpha_c$ & Skewness of colour population \\
		$\sigma_{M_B}$ & Population magnitude scatter \\
		$\sigma_{x_1}$ & Population stretch scatter \\
		$\sigma_{c}$ & Population colour scatter \\
		$\kappa_{0}$ & Extra colour dispersion\\
		$\kappa_{1}$ & Redshift-dependence of extra colour dispersion\\
		\hline
		$m_B$ & True flux\\
		$x_1$ & True stretch \\
		$c$ & True colour \\
		$z$ & True redshift \\
		$M_B$ & Derived absolute magnitude \\
		$\mu$ & Derived distance modulus \\
		\hline
		$\hat{m}_B$ & Measured flux\\
		$\hat{x}_1$ & Measured stretch \\
		$\hat{c}$ & Measured colour \\
		$C$ & Covariance on flux, stretch and colour \\
		$\hat{z}$ & Observed redshift \\
		$\hat{m}$ & Determined mass probability \\
		$\hat{p}$ & Determined outlier probability \\
		\hline
	\end{tabular}
\end{table}

@@

We can thus include into our model parametrised selection effects by including the covariance matrix of selection effect uncertainty. Formally, we include deviations from the determined mean selection function parameters with parameter vector $\Delta S$, and apply a normal prior on this parameter as per the determined uncertainty covariance matrix. {\green Whilst this uncertainty encapsulates the potential error from the simulations not matching the analytic approximations, it does not cover potential variations of the selection function at the top level - varying cosmology or spectroscopic efficiency. Tests with changing the selection efficiency model show that the uncertainty introduced is sub-dominant to the uncertainty introduced from assuming an analytic form.}

@@

In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and our full treatment of systematics. {\green Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016} and shown in Table \ref{tab:dist}. Initial tests were also done with a second, Gaussian population with colour and stretch populations centered on zero and with respective width $0.1$ and $1$, however cosmological parameters were not significantly impacted by choice of the underlying population and we continue using only the {\sk} population for computational efficiency.} The selection effects were quantified by comparing all the generated supernovae to those that pass our simulated cuts, as shown in Figure~\ref{fig:sf_snana}. It is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.3\sigma$ when taking into account the uncertainty on fits to $w$. {\green The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population. As our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and represents a minimum bound on permissible uncertainty values.}}
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ (scatter) & Bias \\
		\hline
		{\gten} Stat + Syst  &  $-1.00\pm 0.10\ (0.08)$ & $0.0\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.95\pm 0.10\ (0.06)$ & $0.5\sigma$  \\
		{\gten}  Stat         &  $-1.00\pm 0.08\ (0.08)$ & $0.0\sigma$   \\
		{\celeven} Stat         &  $-0.95\pm 0.08\ (0.05)$ & $0.6\sigma$   \\
		\hline
	\end{tabular}
\end{table}

@@

We investigate the cosmological bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty reported from fitting the SALT2 model to light curves is incorrect, non-Gaussian and biased. {\green Our current model and techniques can quantify the effect of different scatter models on biasing the observed summary statistics, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model. We note that the incorrect summary statistics do not only bias cosmology, but also cause significant bias with the recovery of the standardisation parameters $\alpha$ and $\beta$. Fixing $\alpha$ and $\beta$ values to simulation values when fitting aids in recovering correct cosmology, but does not remove all bias.}

@@

For the sample size of the DES and LowZ supernova samples (of order 350 supernova), the bias from intrinsic scatter models is sub-dominant to the statistical uncertainty, as shown in Figure \ref{fig:bulk_posterior}. {\green For our full systematics model, the bias represents a deviation between $0\sigma$ to $0.5\sigma$ depending on scatter model, and as such we will leave more complicated treatment of them for future work.}

@@

\citep{Scolnic2017}

{\red Introduce BBC. Highlight its use on Panstars}

{\red have plots for w-w comparison for g10 and c11}

{\red conclude that correlation is good, yay, have extra uncertainty, but this can be explained from previous section}

BHM Mean for G10: -0.998, BBC mean -1.042, average diff -0.044, difference scatter 0.057

BHM Mean for C11: -0.945, BBC mean -0.978, average diff -0.033, difference scatter 0.062

{\red For systematics paper, also find 60\% reduction in uncert using planck priors}

\begin{table}
	\centering
	\caption{Good table right here. Yeah boi.}
	\label{tab:bulk_summary_bbc}
	\begin{tabular}{l|ccc}
		\hline
		Model & {\gten} & {\celeven} & Mean ({\gten} + {\celeven}) \\
		\hline
		Steve $\langle w \rangle$           & $-0.998$ & $-0.945$ & $-1.010$ \\
		BBC $\langle w \rangle$             & $-1.044$ & $-0.978$ & $-0.972$ \\
		(Steve - BBC)  $\langle w \rangle$  & $-0.044$ & $-0.033$ & $-0.039$ \\
		\hline
		(Steve - BBC)  $\sigma_w$           & $0.057$ & $0.062$ & $0.060$ \\
		\hline
	\end{tabular}
\end{table}

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_comparison.pdf}
	\end{center}
	\caption{Whoa.}
	\label{fig:bulk_comparison}
\end{figure}

>>>>

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova colour population's mean and skewness are highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A common solution to this computational cost problem is to bin the supernovae, which then produces a matrix of arbitrary size. Whilst binning data does result in some loss of information, recent works tested against simulations show that this information is not cosmologically significant \citep{Scolnic2016, Kessler2017}.

@@

The diverse approaches and implementations applied to supernova cosmology are a response to the significant challenges and complications faced by when performing supernova cosmology analyses. In this Section, we outline several of the most prominent challenges.

@@

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SNIa, datasets are commonly formed from the discoveries of multiple surveys. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

\begin{table}
	\centering
	\caption{ Parameters defined in our model and a summary of their use. The parameters are broken into multiple sections, the top for parameters which are defined globally across all surveys, the second section for parameters which are defined for each survey, the third section are parameters which are defined for each supernova. The bottom section shows the observables (not parameters) that are the input data to the model.}
	\label{tab:param_summary}
	\begin{tabular}{ll}
		\hline
		Parameter & Description \\
		\hline
		$\Omega_m$  & Matter density  \\
		$w$  &  Dark energy equation of state  \\
		$\alpha$ & Stretch standardisation   \\
		$\beta$         &  Colour standardisation   \\
		$\delta(0)$ & Scale of the mass-magnitude correction\\
		$\delta(\infty)/\delta(0)$ & Redshift-dependence of mass-magnitude correction\\
		$\delta\mathcal{Z}_i$ & Systematics scale\\
		$\delta S$ & Selection effect deviation \\
		$\langle M_B \rangle$ & Mean absolute magnitude \\
		%$s_c$ & Colour bias\\
		\hline
		$\langle x_1^i \rangle$ & Mean stretch nodes\\
		$\langle c^i \rangle$ & Mean colour nodes\\
		$\alpha_c$ & Skewness of colour population \\
		$\sigma_{M_B}$ & Population magnitude scatter \\
		$\sigma_{x_1}$ & Population stretch scatter \\
		$\sigma_{c}$ & Population colour scatter \\
		$\kappa_{0}$ & Extra colour dispersion\\
		$\kappa_{1}$ & Redshift-dependence of extra colour dispersion\\
		\hline
		$m_B$ & True flux\\
		$x_1$ & True stretch \\
		$c$ & True colour \\
		$z$ & True redshift \\
		$M_B$ & Derived absolute magnitude \\
		$\mu$ & Derived distance modulus \\
		\hline
		$\hat{m}_B$ & Measured flux\\
		$\hat{x}_1$ & Measured stretch \\
		$\hat{c}$ & Measured colour \\
		$C$ & Covariance on flux, stretch and colour \\
		$\hat{z}$ & Observed redshift \\
		$\hat{m}$ & Determined mass probability \\
		$\hat{p}$ & Determined outlier probability \\
		\hline
	\end{tabular}
\end{table}

@@

We can thus include into our model parametrised selection effects by including the covariance matrix of selection effect uncertainty. Formally, we include deviations from the determined mean selection function parameters with parameter vector $\Delta S$, and apply a normal prior on this parameter as per the determined uncertainty covariance matrix.  Whilst this uncertainty encapsulates the potential error from the simulations not matching the analytic approximations, it does not cover potential variations of the selection function at the top level - varying cosmology or spectroscopic efficiency. Tests with changing the intrinsic scatter model used in the selection efficiency simulations show that the uncertainty introduced is negligible.

@@

In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and our full treatment of systematics. Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016} and shown in Table \ref{tab:dist}. Initial tests were also done with a second, Gaussian population with colour and stretch populations centered on zero and with respective width $0.1$ and $1$, however cosmological parameters were not significantly impacted by choice of the underlying population and we continue using only the {\sk} population for computational efficiency. The selection effects were quantified by comparing all the generated supernovae to those that pass our simulated cuts, as shown in Figure~\ref{fig:sf_snana}. It is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.5\sigma$ when taking into account the uncertainty on fits to $w$. The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population. As our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and represents a minimum bound on permissible uncertainty values.}
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ (scatter) & Bias \\
		\hline
		{\gten} Stat + Syst  &  $-1.00\pm 0.10\ (0.08)$ & $0.0\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.95\pm 0.10\ (0.06)$ & $0.5\sigma$  \\
		{\gten}  Stat         &  $-1.00\pm 0.08\ (0.08)$ & $0.0\sigma$   \\
		{\celeven} Stat         &  $-0.95\pm 0.08\ (0.05)$ & $0.6\sigma$   \\
		\hline
	\end{tabular}
\end{table}

@@

We investigate the cosmological bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. We then also test a set of 100 simulations generated using only coherent magnitude scatter, and also find $w=-1.00$. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty reported from fitting the SALT2 model to light curves is incorrect, non-Gaussian and biased. Our current model and techniques can quantify the effect of different scatter models on biasing the observed summary statistics, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model.

@@

For the sample size of the DES and LowZ supernova samples (of order 350 supernova), the bias from intrinsic scatter models is sub-dominant to the statistical uncertainty, as shown in Figure \ref{fig:bulk_posterior}. For our full systematics model, the bias represents a deviation between $0\sigma$ to $0.5\sigma$ depending on scatter model, and as such we will leave more complicated treatment of them for future work.

@@

We compare the results of our model against those of the BBC method \citep{Kessler2017}. BBC has been used in prior analyses, such as the Pantheon sample analysis of \citet{Scolnic2017}, as a leading supernova cosmology method provides a good consistency as to the current levels of accuracy in recovering cosmological parameters.

To this end, we take the results of the BBC method also run on our set of 200 validation simulations and compare the recovered $w$ values to those of our method. The results are detailed in Table \ref{tab:bulk_summary_bbc}, and a scatter plot of the simulation results is presented in Figure \ref{fig:bulk_comparison}. These show that the difference in recovered $w$ over different intrinsic scatter models is comparable, with our methodology having a small (but not statistically significant) reduction in the sensitivity to intrinsic scatter model. However, as the BBC method introduces a bias to lower $w$ for {\gten} and a bias towards higher $w$ for {\celeven}, taking the average result over both {\gten} and {\celeven} only results in a small bias of $0.01$ in $w$. As our method recovers $w$ without bias for {\gten} and finds a biased high $w$ for {\celeven}, the average bias over the two scatter models is $-0.028$. When comparing both the {\gten} and {\celeven} set of simulations independently, our model differs from BBC in its average prediction of $w$ by $-0.044$ and $-0.033$ respectively. For the {\gten} model this difference is a result of bias in the BBC results, however for the {\celeven} simulations this is a result of both bias from BBC, and a larger bias from our method. These results also allow us to state the expectation values for $w$ when run on the DES three year spectroscopic survey sample -- factoring in the calculated uncertainty from Planck priors (shrinking the scatter from $0.06$ to $0.04$), we expect our BHM method to, on average, recover $w_{\rm BHM} = w_{\rm BBC} -0.04\pm0.04$.

Having established that existing methodologies exhibit similar shifts in their recovery of $w$ compared to our method, future work will focus on improving the parametrisation of intrinsic scatter model into our framework, with the goal of minimising the effect of the underlying scatter model on the recovery of cosmology.

\begin{table}
	\centering
	\caption{Characterising the bias on $w$ using the 100 simulations for the {\gten} scatter model and 100 simulations for {\celeven} scatter model. The mean $w$ value for our method and BBC are presented, along with the mean when averaging the difference between our method and BBC for each individual simulation. Finally, we also characterise the scatter between the methods in the final row.}
	\label{tab:bulk_summary_bbc}
	\begin{tabular}{l|ccc}
		\hline
		Model & {\gten} & {\celeven} & Mean ({\gten} + {\celeven}) \\
		\hline
		Steve $\langle w \rangle$           & $-0.998$ & $-0.945$ & $-1.010$ \\
		BBC $\langle w \rangle$             & $-1.044$ & $-0.978$ & $-0.972$ \\
		(Steve - BBC)  $\langle w \rangle$  & $-0.044$ & $-0.033$ & $-0.039$ \\
		\hline
		(Steve - BBC)  $\sigma_w$           & $0.057$ & $0.062$ & $0.060$ \\
		\hline
	\end{tabular}
\end{table}

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_comparison.pdf}
	\end{center}
	\caption{Recovered $w$ for the 200 validation simulations with full treatment of statistical and systematic errors. Uncertainty on the recovered $w$ value is shown for every second data point for visual clarity.}
	\label{fig:bulk_comparison}
\end{figure}
